---
title: "Intoxication Detection Project"
author: "Norman, Elsa; McCartney, Connor; Dertke, Mike"
date: 'Last Updated: Fri Dec 12, 2025' 
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**Using smartphone accelerometer data to detect user intoxication**

***This project uses a dataset from the UC Irvine Machine Learning Repository at https://archive.ics.uci.edu/dataset/515/bar+crawl+detecting+heavy+drinking. The dataset contains smartphone accelerometer data and TAC measurements for 13 subjects while participating in the Ohio State University Annual Senior Bar Crawl. This project investigates whether smartphone accelerometer data can be used to detect intoxication for the 13 subjects in this study.***



```{css solution-region, echo=FALSE}
.solution {
    background-color: #232D4B10;
    border-style: solid;
    border-color: #232D4B;
    padding: .5em;
    margin: 20px
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## [Models]{style="color:#FF7F50"}

We need to first load the necessary libraries. 
```{r message=FALSE}
library(randomForest)
library(gbm)
library(tree)
library(e1071)
library(class)
library(ggplot2)
```

Load in dataset and split into testing and training values
```{r}
set.seed(52)
data = read.csv("bar+crawl+detecting+heavy+drinking\\data\\average_accel.csv")

data$label = as.factor(data$label)

# rerrange rows to ensure representative sampling
data = data[sample(nrow(data)), ]

data_rows = floor(0.80 * nrow(data))
train_indices = sample(c(1:nrow(data)), data_rows)
train_data = data[train_indices,]
test_data = data[-train_indices,]
```

```{r}
accuracies = rep(NA, 8)
precisions = rep(NA, 8)
recalls = rep(NA, 8)
f1s = rep(NA, 8)
```


### [1. Logistic Regression Model]{style="color:#FF7F50"}

Run the logistic regression model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
model = glm(label ~ ., data = train_data, family = binomial)

test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))

pred_probs = predict(model, test_data, type = "response")
pred = factor(ifelse(pred_probs < 0.5, "sober", "intoxicated"))


tp = sum(test_labels == pred & pred == "intoxicated")
fp = sum(test_labels != pred & pred == "intoxicated")
tn = sum(test_labels == pred & pred == "sober")
fn = sum(test_labels != pred & pred == "sober")

accuracy = mean(test_labels == pred); accuracies[1] = accuracy
precision = tp / (tp + fp); precisions[1] = precision
recall = tp / (tp + fn); recalls[1] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[1] = f1

cat(noquote(paste0("Logistic Regression\n\nAccuracy: ", round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4), "\n\n")))

conf_matrix = table(Predicted = pred, True = test_labels)
cm_swapped = conf_matrix[c(2, 1), ] # swap rows
cm_swapped
```


### [2. Bagging Model]{style="color:#FF7F50"}

First conduct parameter tuning on mtry, then run the bagging model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
bag.acc.values = rep(NA, 4)

for (i in 1:4) {
  set.seed(12)
  bag = randomForest(label ~ ., data = train_data, mtry = i, importance = TRUE)
  pred = predict(bag, test_data, type="response")
  test_labels = test_data$label
  bag.acc.values[i] = mean(test_labels == pred)
}

mtry = which.max(bag.acc.values)

names(bag.acc.values) = c(1,2,3,4)
bag.acc.values
```

```{r}
set.seed(12)
bag = randomForest(label ~ ., data = train_data, mtry = mtry, importance = TRUE)
bag

importance(bag)
varImpPlot(bag)

# test accuracy
pred = predict(bag, test_data, type="response")


test_labels = test_data$label

tp = sum(test_labels == pred & pred == 1)
fp = sum(test_labels != pred & pred == 1)
tn = sum(test_labels == pred & pred == 0)
fn = sum(test_labels != pred & pred == 0)

accuracy = mean(test_labels == pred); accuracies[2] = accuracy
precision = tp / (tp + fp); precisions[2] = precision
recall = tp / (tp + fn); recalls[2] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[2] = f1

cat(noquote(paste0("Bagging\n\nAccuracy: ", round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
pred_labels = factor(pred, labels = c("sober", "intoxicated"))

conf_matrix = table(Predicted = pred_labels, True = test_labels)
conf_matrix
```


### [3. Random Forest Model]{style="color:#FF7F50"}

Run the random forest model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
set.seed(12)
rf = randomForest(label ~ ., data = train_data, importance = TRUE)
rf

importance(rf)
varImpPlot(rf)

# test accuracy
pred = predict(rf, test_data, type="response")


test_labels = test_data$label

tp = sum(test_labels == pred & pred == 1)
fp = sum(test_labels != pred & pred == 1)
tn = sum(test_labels == pred & pred == 0)
fn = sum(test_labels != pred & pred == 0)

accuracy = mean(test_labels == pred); accuracies[3] = accuracy
precision = tp / (tp + fp); precisions[3] = precision
recall = tp / (tp + fn); recalls[3] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[3] = f1

cat(noquote(paste0("Random Forest\n\nAccuracy: ", round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
pred_labels = factor(pred, labels = c("sober", "intoxicated"))

conf_matrix = table(Predicted = pred_labels, True = test_labels)
conf_matrix
```


### [4. Boosting Model]{style="color:#FF7F50"}

First conduct parameter tuning on interaction.depth and cv.folds, then run the bagging model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
depth.acc.values = rep(NA, 4)
folds.acc.values = rep(NA, 7)

for (i in 1:4) {
  set.seed(12)
  
  boost = gbm(label ~ ., data = train_data, distribution = "gaussian", n.trees = 150, interaction.depth = i, cv.folds = 3)
  
  best_iter = gbm.perf(boost, method = "cv")
  pred = predict(boost, test_data, n.trees = best_iter, type = "response")

  test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
  pred_labels = factor(ifelse(pred < 1.5, "sober", "intoxicated"))
  
  depth.acc.values[i] = mean(test_labels == pred_labels)
}
depth = which.max(depth.acc.values)

names(depth.acc.values) = c(1,2,3,4)
depth.acc.values

for (i in 2:7) {
  boost = gbm(label ~ ., data = train_data, distribution = "gaussian", n.trees = 150, interaction.depth = depth, cv.folds = i)
  
  best_iter = gbm.perf(boost, method = "cv")
  pred = predict(boost, test_data, n.trees = best_iter, type = "response")

  test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
  pred_labels = factor(ifelse(pred < 1.5, "sober", "intoxicated"))
  
  folds.acc.values[i] = mean(test_labels == pred_labels)
}
folds = which.max(folds.acc.values)

names(folds.acc.values) = c(1:7)
folds.acc.values
```

```{r}
set.seed(12)

boost = gbm(label ~ ., data = train_data, distribution = "gaussian", n.trees = 150, interaction.depth = depth, cv.folds = folds)
boost

summary(boost)

# test accuracy
best_iter = gbm.perf(boost, method = "cv")
pred = predict(boost, test_data, n.trees = best_iter, type = "response")

test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
pred_labels = factor(ifelse(pred < 1.5, "sober", "intoxicated"))


tp = sum(test_labels == pred_labels & pred_labels == "intoxicated")
fp = sum(test_labels != pred_labels & pred_labels == "intoxicated")
tn = sum(test_labels == pred_labels & pred_labels == "sober")
fn = sum(test_labels != pred_labels & pred_labels == "sober")

accuracy = mean(test_labels == pred_labels); accuracies[4] = accuracy
precision = tp / (tp + fp); precisions[4] = precision
recall = tp / (tp + fn); recalls[4] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[4] = f1

cat(noquote(paste0("Boosting\n\nAccuracy: ", round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

conf_matrix = table(Predicted = pred_labels, True = test_labels)
cm_swapped = conf_matrix[c(2, 1), ] # swap rows
cm_swapped
```

### [5. Classification Tree Model]{style="color:#FF7F50"}

Run the classification tree model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
tree = tree(label ~ ., data = train_data)

summary(tree)

plot(tree)
text(tree)

pred = predict(tree, test_data, type = "class")

# test accuracy
test_labels = test_data$label

tp = sum(test_labels == pred & pred == 1)
fp = sum(test_labels != pred & pred == 1)
tn = sum(test_labels == pred & pred == 0)
fn = sum(test_labels != pred & pred == 0)

accuracy = mean(test_data$label == pred); accuracies[5] = accuracy
precision = tp / (tp + fp); precisions[5] = precision
recall = tp / (tp + fn); recalls[5] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[5] = f1

cat(noquote(paste0("Classification Tree\n\nAccuracy: ", round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
pred_labels = factor(pred, labels = c("sober", "intoxicated"))

conf_matrix = table(Predicted = pred_labels, True = test_labels)
conf_matrix
```


### [6. Support Vector Machine Model]{style="color:#FF7F50"}

First conduct parameter tuning on kernel, then run the SVM model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
kernel.options = c("polynomial", "radial", "sigmoid")
kernel.acc.values = rep(NA, 3)

for (i in 1:3) {
  svm.model = svm(label ~ ., data = train_data, kernel = kernel.options[i])
  
  pred = predict(svm.model, test_data, type="response")
  test_labels = test_data$label
  
  kernel.acc.values[i] = mean(test_labels == pred)
}
kernel = kernel.options[which.max(kernel.acc.values)]

names(kernel.acc.values) = kernel.options
kernel.acc.values
```

```{r}
svm.model = svm(label ~ ., data = train_data, kernel = kernel)

# test accuracy
pred = predict(svm.model, test_data, type="response")


test_labels = test_data$label

tp = sum(test_labels == pred & pred == 1)
fp = sum(test_labels != pred & pred == 1)
tn = sum(test_labels == pred & pred == 0)
fn = sum(test_labels != pred & pred == 0)

accuracy = mean(test_labels == pred); accuracies[6] = accuracy
precision = tp / (tp + fp); precisions[6] = precision
recall = tp / (tp + fn); recalls[6] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[6] = f1

cat(noquote(paste0("Support Vector Machine\n\nAccuracy: ", 
                   round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

test_labels = factor(test_data$label, labels = c("sober", "intoxicated"))
pred_labels = factor(pred, labels = c("sober", "intoxicated"))

conf_matrix = table(Predicted = pred_labels, True = test_labels)
conf_matrix
```

### [7. Hierarchical Clustering]{style="color:#FF7F50"}

First conduct parameter tuning on linkage method, then run the hierarchical clustering model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
linkage.options = c("single", "complete", "average")
linkage.acc.values = rep(NA, 3)
linkage.prec.values = rep(NA, 3)
linkage.rec.values = rep(NA, 3)
linkage.f1.values = rep(NA, 3)

for (i in 1:3) {
  hc = hclust(dist(data), method = linkage.options[i])
  
  pred = cutree(hc, 2)
  pred = data.frame(predictions = pred)
  
  train_labels = factor(data$label, labels = c("sober", "intoxicated"))
  pred_labels = factor(ifelse(pred < 1.5, "sober", "intoxicated"))
  
  tp = sum(train_labels == pred_labels & pred_labels == "intoxicated")
  fp = sum(train_labels != pred_labels & pred_labels == "intoxicated")
  tn = sum(train_labels == pred_labels & pred_labels == "sober")

  linkage.acc.values[i] = mean(train_labels == pred_labels)
  linkage.prec.values[i] = tp / (tp + fp) #precision
  linkage.rec.values[i] = tp / (tp + fn)
  linkage.f1.values[i] = 2 * (precision * recall) / (precision + recall)
}

linkage = linkage.options[which.max(linkage.prec.values)]
#accuracies
names(linkage.acc.values) = linkage.options
linkage.acc.values
#precision values
names(linkage.prec.values) = linkage.options
linkage.prec.values
#recall values
names(linkage.rec.values) = linkage.options
linkage.rec.values
#f1 values
names(linkage.f1.values) = linkage.options
linkage.f1.values
```

```{r}
hc.model = hclust(dist(data), method = linkage)

pred = cutree(hc.model, 2)
pred = data.frame(predictions = pred)


train_labels = factor(data$label, labels = c("sober", "intoxicated"))
pred_labels = factor(ifelse(pred < 1.5, "sober", "intoxicated"))


tp = sum(train_labels == pred_labels & pred_labels == "intoxicated")
fp = sum(train_labels != pred_labels & pred_labels == "intoxicated")
tn = sum(train_labels == pred_labels & pred_labels == "sober")
fn = sum(train_labels != pred_labels & pred_labels == "sober")

accuracy = mean(train_labels == pred_labels); accuracies[7] = accuracy
precision = tp / (tp + fp); precisions[7] = precision
recall = tp / (tp + fn); recalls[7] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[7] = f1

cat(noquote(paste0("Hierarchical Clustering (single linkage)\n\nAccuracy: ",
                   round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

conf_matrix = table(Predicted = pred_labels, True = train_labels)
cm_swapped = conf_matrix[c(2, 1), ] # swap rows
cm_swapped
```


### [8. K-Nearest Neighbors Model]{style="color:#FF7F50"}

First conduct parameter tuning on k, then run the KNN model and obtain accuracy, precision, recall, and F1-score.

#### {.solution}
```{r}
set.seed(12)

normalize = function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

data_norm = as.data.frame(lapply(data[2:5], normalize))

train_data = data_norm[train_indices, ]
test_data  = data_norm[-train_indices, ]

train_labels = data$label[train_indices]
test_labels  = data$label[-train_indices]


k.options = c(3,5,7,9,11,13,15,17)

k.acc.values = rep(NA, 8)
for (i in 1:8) {
  set.seed(12)
  
  pred = knn(train = train_data, test = test_data, cl = train_labels, k = k.options[i])
  
  k.acc.values[i] = mean(pred == test_labels)
}
k = k.options[which.max(k.acc.values)]

names(k.acc.values) = k.options
k.acc.values
```

```{r}
set.seed (12)

pred = knn(train = train_data, test = test_data, cl = train_labels, k = k)


tp = sum(test_labels == pred & pred == 1)
fp = sum(test_labels != pred & pred == 1)
tn = sum(test_labels == pred & pred == 0)
fn = sum(test_labels != pred & pred == 0)

accuracy = mean(pred == test_labels); accuracies[8] = accuracy
precision = tp / (tp + fp); precisions[8] = precision
recall = tp / (tp + fn); recalls[8] = recall
f1 = 2 * (precision * recall) / (precision + recall); f1s[8] = f1

cat(noquote(paste0("K-Nearest Neighbors\n\nAccuracy: ", round(accuracy,4),
                   "\nPrecision: ", round(precision,4),
                   "\nRecall: ", round(recall,4),
                   "\nF1: ", round(f1,4))))

test_labels = factor(test_labels, labels = c("sober", "intoxicated"))
pred_labels = factor(pred, labels = c("sober", "intoxicated"))

conf_matrix = table(Predicted = pred_labels, True = test_labels)
conf_matrix
```



### [Analysis]{style="color:#FF7F50"}

#### {.solution}
```{r}
labels = c("Logistic Regression", "Bagging", "Random Forest", "Boosting", "Classification Tree", "SVM", "Hierarchical Clustering", "KNN")

labels = c("LR", "Bag", "RF", "Boost", "Tree", "SVM", "HC", "KNN")

names(accuracies) = labels
names(precisions) = labels
names(recalls) = labels
names(f1s) = labels

cat("Accuracy\n")
round(accuracies,4)
cat("\nPrecision\n")
round(precisions,4)
cat("\nRecall\n")
round(recalls,4)
cat("\nF1\n")
round(f1s,4)

```

```{r}
accuracies = round(accuracies,4) * 100

df = data.frame(Model = labels, Accuracy = accuracies)
df_sorted <- df[order(df$Accuracy), ]
df_sorted$Model <- factor(df_sorted$Model, levels = df_sorted$Model)
```

```{r}
ggplot(df_sorted, aes(x = Model, y = Accuracy)) +
  geom_col(width = 0.6, fill = "steelblue") +  
  geom_text(aes(label = paste0(Accuracy, "%")), 
            vjust = -0.5, size = 4) + 
  scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +  
  labs(title = "Model Accuracy Comparison",
       x = "Model",
       y = "Accuracy (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```